preciseTAD: A machine learning framework for precise 3D domain boundary prediction at base-level resolution

### 2020-09-17 Manuscript/manuscript.Rmd
Results finished

When trained using Arrowhead and Peakachu ground truth boundaries at 5 kb and 10 kb, respectively, the *preciseTAD* model predicted a total of 12,258 TAD and 15,707 chromatin loop boundaries in GM12878, as well as 9,603 TAD and 11,154 chromatin loop boundaries in K562 cell line (Supplementary Table 4). We reported less predicted TAD boundaries at 5 kb than Arrowhead on both cell lines (Table 1, Supplementary Table 3). This can be attributed to Arrowhead's inflation of called TADs, that, when visualized, often do not correspond to domains enriched in internal interactions (Figure 5A) and signal of known drivers of domain boundaries (Figure 5B). *preciseTAD* also predicted fewer chromatin loop boundaries than Peakachu (Table 1, Supplementary Table 3). This can be attributed to the fact that *preciseTAD* leverages four known drivers of 3D chromatin, including CTCF, RAD21, SMC3, and ZNF143. This is in contrast to Peakachu, which considers only CTCF sites to mark the predicted boundaries as true positives. These results suggest that genome annotation-driven predicted boundaries may be more biologically relevant than those detected from Hi-C data alone.

![**preciseTAD boundaries are enriched for known molecular drivers of 3D chromatin.** (A) The location of Arrowhead-called TAD boundaries (blue) vs. *preciseTAD*-predicted TAD boundaries (green) on GM12878 data (chr14:50085000-50800000). The black line represents the predicted probability of each base being a TAD boundary. (B) A zoomed-in portion of the genome shows the preciseTAD boundary region (PTBR), a yellow-highlighted cluster of bases with high probability of being a boundary. Underneath is the genome browser visualization of bigwig files for known molecular drivers of 3D chromatin including CTCF, RAD21, SMC3, and ZNF143 associated with the PTBR. (C, D) Profile plots comparing the signal strength of CTCF, RAD21, SMC3, and ZNF143 around Arrowhead-called boundaries (blue, C), Peakachu loop boundaries (red, D) vs. *preciseTAD*-predicted boundaries (green).](Figures/figure5.png)

Next, we evaluated the biological significance of *preciseTAD* boundaries. Peak regions associated with known molecular drivers of 3D chromatin, including CTCF, RAD21, SMC3, and ZNF143, colocalized more frequently around *preciseTAD*-predicted TAD and loop boundaries compared to Arrowhead-called TAD and Peakachu loop boundaries respectively (Figure 5C, 5D; Supplementary Figure 5A, 5B). Surprised by the poor signal distribution around Arrowhead boundaries, we compared signals centered on boundaries called by Arrowhead, Peakachu, and our SpectralTAD [@Cresswell:2020ab] TAD caller. We confirmed the poor signal distribution around Arrowhead boundaries, in contrast to the well-performing Peakachu- and SpectralTAD-called boundaries (Supplementary Figure 5C, D). Signal heatmaps indicated that *preciseTAD*-predicted boundaries were more enriched for the same genomic annotations than either Arrowhead or Peakachu boundaries alone (Supplementary Figures 6-7). *preciseTAD* boundaries were statistically significantly closer to the top-ranked TFBSs (Wilcoxon p-value < 0.001 versus Arrowhead and Peakachu boundaries, respectively, Figure 6A, Supplementary Figure 8). These results indicate that *preciseTAD*-predicted boundaries better reflected the known biology of boundary formation.

@@

Previous studies suggest that TAD boundaries are conserved across cell lines [@dixon2012topological; @krefting2018evolutionary; @nora2012spatial; @sexton2012three]. To assess the level of cross-cell-line conservation, we evaluated the overlap between cell line-specific boundaries detected by *preciseTAD*, Arrowhead, and Peakachu. Boundaries involving Arrowhead were flanked by 5 kb, while boundaries involving Peakachu were flanked by 10 kb. Only 26% and 49% of boundaries were conserved between cell lines for Arrowhead and Peakachu boundaries (J=0.186 and J=0.388), respectively (Figure 6D, 6D. However, we found 45%/56% of *preciseTAD*-predicted domain boundaries using models trained on Arrowhead/Peakachu data (J=0.383 and J=0.444), respectively, were conserved between GM12878 and K562 cell lines (Figure 6C, 6E). The better conservation of *preciseTAD*-predicted boundaries further supports the notion of their higher biological relevance.

![**preciseTAD-predicted boundaries are both statistically significantly closer to CTCF sites and more conserved across cell lines**. (A) Log2 genomic distances from called and predicted boundaries to nearest CTCF sites. The p-values are from the Wilcoxon Rank Sum test. (B) Venn diagrams illustrating the levels of conservation between domain boundaries for GM12878 (red) and K562 (blue) cell lines identified by Arrowhead boundaries themselves, Arrowhead-trained *preciseTAD* models (C), Peakachu boundaries themselves (C), and Peakachu-trained *preciseTAD* model (D). J - the Jaccard Index evaluating the similarity between each set of boundaries flanked by resolution.](Figures/figure6.png)

@@

The majority of boundaries predicted by the Arrowhead-trained *preciseTAD* model represented a subset of the larger group of boundaries predicted by the Peakachu-trained model. We found that 88.8% and 95.8% of our predicted TAD boundaries were overlapped by predicted loop boundaries for GM12878 and K562, respectively (Supplementary Figure 9). This is expected as loop boundaries (Peakachu) are more abundant, while TAD boundaries (Arrowhead) represent the narrower level of the 3D chromatin organization. Together with the results in Figure 4, the high overlap between boundaries predicted by Arrowhead- and Peakachu-trained models suggests that TAD and loop boundaries may be driven by similar molecular mechanisms. Guided by the high predictive performance when training and testing on different cell lines (Figure 4D), we opted to evaluate whether *preciseTAD* could be leveraged to predict boundaries using unseen base level cell line-specific annotation data. To do so, we evaluated two separate scenarios for each type of ground truth boundary: 1) training on GM12878 and predicting boundaries on GM12878 (GM on GM) vs training on K562 and predicting on GM12878 (K on GM), and 2) training on K562 and predicting boundaries on K562 (K on K) vs training on GM12878 and predicting boundaries on K562 (GM on K). Using Arrowhead ground truth, 76% (J=0.701) and 81% (J=0.751) of predicted boundaries were overlapped in each scenario, respectively (Supplementary Figure 10A, 10B). Likewise, when using Peakachu ground truth, there was 85% (J=0.705) and 88% (J=0.759) of overlap between predicted boundaries, respectively (Supplementary 10C, 10D). Furthermore, boundaries predicted on unseen annotation data exhibited similar levels of enrichment for CTCF, RAD21, SMC3, and ZNF143 as did those trained and predicted on the same cell line. These results open up the possibility of using preciseTAD to predict TAD or loop boundaries for cell lines that do not currently have Hi-C data.

@@

**Supplementary Table 4.** Genomic coordinates of *preciseTAD*-predicted boundary regions (PTBR) and points (PTBP) for GM12878 and K562, using Arrowhead TAD and Peakachu chromatin loop boundary coordinates as ground truth. For PTBRs, the start and end coordinates define the clusters of spatially proximal bases with the probability of being a boundary equal to 1 and $\epsilon=10000$. For PTBPs, the start and end (start+1) coordinates define the uniquely identified boundary coordinate within each PTBR. All coordinates are in hg19/GRCh37 reference genome. (Work in progress)

**Supplementary Figure 5.** Profile plots comparing the signal strength of top TFBS around flanked (A) Arrowhead called TAD boundaries (blue) and *preciseTAD*-predicted TAD boundaries (green), (B) Peakachu chromatin loop boundaries (red) and preciseTAD predicted loop boundaries (green) on K562, (C) Arrowhead called TAD boundaries (blue), Peakachu chromatin loop boundaries (red), and SpectralTAD called TAD boundaries (green) on GM12878 and (D) on K562.

**Supplementary Figure 6. *preciseTAD* boundaries are more enriched for known molecular drivers of 3D chromatin.** Enriched heatmaps comparing the enrichment of Arrowhead called TAD boundaries vs *preciseTAD*-predicted TAD boundaries around key transcription factor binding sites known to contribute to the 3D chromatin architecture for (A) GM12878 and (B) K562.

**Supplementary Figure 7. *preciseTAD* boundaries are more enriched for known molecular drivers of 3D chromatin.** Enriched heatmaps comparing the enrichment of Peakachu chromatin loop boundaries vs *preciseTAD*-predicted chromatin boundaries around key transcription factor binding sites known to contribute to the 3D chromatin architecture for (A) GM12878 and (B) K562.

**Supplementary Figure 8. *preciseTAD* boundaries are spatially closer to known molecular drivers of 3D chromatin.** Boxplots comparing the log2 genomic distances from predicted and called boundaries to the nearest (A) GM12878-specific and (B) K562-specific CTCF, RAD21, SMC3, and ZNF143 transcription factor binding sites. The p-value is derived from the Wilcoxon Rank Sum test.

**Supplementary Figure 9.*preciseTAD*-predicted TAD boundaries are nested within *preciseTAD*-predicted chromatin loop boundaries.** Venn diagrams illustrating the agreement between boundaries predicted using Arrowhead- and Peakachu-trained models for (A) GM12878 and (B) K562. Domain boundaries were flanked by resolution (5 kb for models using Arrowhead and 10 kb for Peakachu).

**Supplementary Figure 10. *preciseTAD* accurately predicts boundaries on cell lines using unseen annotation data.** Venn diagrams comparing flanked predicted boundaries using Arrowhead ground truth to: (A) train on GM12878 and predict on GM12878 (GM on GM) vs train on K562 and predict on GM12878 (K on GM), (B) train on K562 predict on K562 (K on K) vs train on GM12878 predict on K562 (GM on K). (C-D) Using Peakachu ground truth. The corresponding profilePlots (deepTools version 2.0) are provided to the right. Predicted boundaries using Arrowhead ground truth were flanked by 5 kb, while boundaries predicted using Peakachu ground truth were flanked by 10 kb.

>>>>

When trained using Arrowhead and Peakachu ground truth boundaries at 5 kb and 10 kb, respectively, the *preciseTAD* model predicted a total of 12,258 TAD and 15,707 chromatin loop boundaries in GM12878, as well as 9,603 TAD and 11,154 chromatin loop boundaries in K562 cell line (Supplementary Table 4). We reported less predicted TAD boundaries at 5 kb than Arrowhead on both cell lines (Table 1, Supplementary Table 3). This can be attributed to Arrowhead's inflation of called TADs, that, when visualized, often do not correspond to domains enriched in internal interactions (Figure 5A) and signal of known drivers of domain boundaries (Figure 5B). *preciseTAD* also predicted fewer chromatin loop boundaries than Peakachu (Table 1, Supplementary Table 3). This can be attributed to Peakachu's use of only CTCF sites to call boundaries, while *preciseTAD* leverages four known drivers of 3D chromatin, including CTCF, RAD21, SMC3, and ZNF143. These results suggest that genome annotation-driven predicted boundaries may be more biologically relevant than those detected from Hi-C data alone.

![**preciseTAD boundaries are enriched for known molecular drivers of 3D chromatin.** (A) The location of Arrowhead-called TAD boundaries (blue) vs. *preciseTAD*-predicted TAD boundaries (green) on GM12878 data (chr14:50085000-50800000). The black line represents the predicted probability of each base being a TAD boundary. (B) A zoomed-in portion of the genome shows the preciseTAD boundary region (PTBR, highlighted yellow), a cluster of bases with high probability of being a boundary, and the corresponding signal profiles of CTCF, RAD21, SMC3, and ZNF143. (C, D) Signal profile plots comparing the strength of CTCF, RAD21, SMC3, and ZNF143 binding around Arrowhead-called boundaries (blue, C), Peakachu loop boundaries (red, D) vs. *preciseTAD*-predicted boundaries (green).](Figures/figure5.png)

Next, we evaluated the biological significance of *preciseTAD* boundary points. Signal of four known molecular drivers of 3D chromatin (CTCF, RAD21, SMC3, and ZNF143) colocalized more frequently around *preciseTAD*-predicted boundary points, as compared to Arrowhead-called TAD and Peakachu loop boundaries, respectively (Figure 5C, 5D; Supplementary Figure 5A, 5B). Surprised by the poor signal distribution around Arrowhead boundaries, we compared signals centered on boundaries called by Arrowhead, Peakachu, and our SpectralTAD [@Cresswell:2020ab] TAD caller. We confirmed the poor signal distribution around Arrowhead boundaries, in contrast to the relatively well-performing Peakachu- and SpectralTAD-called boundaries (Supplementary Figure 5C, D). Signal enrichment heatmaps confirmed that *preciseTAD*-predicted boundaries were more enriched for the same genomic annotations than either Arrowhead or Peakachu boundaries alone (Supplementary Figures 6-7). *preciseTAD* boundaries were statistically significantly closer to the top-ranked TFBSs (Wilcoxon p-value < 0.001 versus Arrowhead and Peakachu boundaries, respectively, Figure 6A, Supplementary Figure 8). These results indicate that *preciseTAD*-predicted boundaries better reflected the known biology of boundary formation.

@@

Previous studies suggest that TAD boundaries are conserved across cell lines [@dixon2012topological; @krefting2018evolutionary; @nora2012spatial; @sexton2012three]. To assess the level of cross-cell-line conservation, we evaluated the overlap between cell line-specific boundaries detected by *preciseTAD*, Arrowhead, and Peakachu. Only 26% and 49% of boundaries were conserved between cell lines for Arrowhead and Peakachu boundaries (J=0.186 and J=0.388), respectively (Figure 6D, 6D. However, 45%/56% of *preciseTAD*-predicted domain boundaries were conserved between GM12878 and K562 cell lines when using models trained on Arrowhead/Peakachu data (J=0.383 and J=0.444, respectively, Figure 6C, 6E). The better conservation of *preciseTAD*-predicted boundaries further supports the notion of their higher biological relevance.

![**preciseTAD-predicted boundaries are closer to CTCF sites and more conserved across cell lines**. (A) Log2 genomic distance distribution from called and predicted boundaries to nearest CTCF sites. The p-values are from the Wilcoxon Rank Sum test. (B-E) Venn diagrams illustrating the levels of conservation (overlap) between domain boundaries for GM12878 (red) and K562 (blue) cell lines identified by Arrowhead (B), Peakachu (D), *preciseTAD*-predicted boundaries using Arrowhead- and Peakachu-trained models () Arrowhead-trained *preciseTAD* models (C), Peakachu boundaries themselves (C, E, respectively). Boundaries involving Arrowhead/Peakachu were flanked by 5 kb/10 kb, respectively.](Figures/figure6.png)

@@

The majority of boundaries predicted by the Arrowhead-trained *preciseTAD* model represented a subset of the larger group of boundaries predicted by the Peakachu-trained model. We found that 88.8% and 95.8% [??? These numbers are different from what is shown on Supp Fig 9] of our predicted TAD boundaries were overlapped by predicted loop boundaries for GM12878 and K562, respectively (Supplementary Figure 9). [??? After clarification, MD to rewrite the first two sentences, unclear] This is expected as loop boundaries (Peakachu) are more abundant, while TAD boundaries (Arrowhead) represent the higher level of the 3D chromatin organization. Together with the results in Figure 4C, the high overlap between boundaries predicted by Arrowhead- and Peakachu-trained models suggests that TAD and loop boundaries may be driven by similar molecular mechanisms.

Guided by the high predictive performance when training and testing on different cell lines (Figure 4D), we opted to evaluate whether models trained using Arrowhead/Peakachu ground truth data in one cell line could be leveraged to predict boundaries using annotation data from another cell line. We evaluated two scenarios: 1) training on GM12878 and predicting boundaries on GM12878 (GM on GM) vs. training on K562 and predicting on GM12878 (K on GM), and 2) training on K562 and predicting boundaries on K562 (K on K) vs. training on GM12878 and predicting boundaries on K562 (GM on K). Using Arrowhead-trained models, 76% (J=0.701) and 81% (J=0.751) of predicted boundaries overlapped in both cross-cell-line prediction scenarios (Supplementary Figure 10A, 10B). Likewise, when using Peakachu-trained models, we observed 85% (J=0.705) and 88% (J=0.759) overlaps (Supplementary 10C, 10D). Furthermore, boundaries predicted on unseen annotation data exhibited similar levels of enrichment for CTCF, RAD21, SMC3, and ZNF143 as did those trained and predicted on the same cell line. [??? Where is it shown?] These results indicate that *preciseTAD* pre-trained models can be successfully used to predict domain boundaries for cell lines lacking Hi-C data but for which genome annotation data is available.

@@

**Supplementary Table 4.** hg19/GRCh37 genomic coordinates of *preciseTAD*-predicted boundary regions (PTBR) and points (PTBP) for GM12878 and K562 cell lines, using models trained on Arrowhead TAD and Peakachu chromatin loop boundaries as ground truth. For PTBRs, the start and end coordinates define the clusters of spatially proximal bases with the probability of being a boundary equal to 1. For PTBPs, the start and end (start+1) coordinates define the most likely boundary point within each PTBR.

**Supplementary Figure 5.** Signal profile plots comparing the binding strength of top TFBSs around flanked (A) Arrowhead called TAD boundaries (blue) and *preciseTAD*-predicted TAD boundaries (green) on K562, (B) Peakachu chromatin loop boundaries (red) and preciseTAD predicted loop boundaries (green) on K562, (C) Arrowhead called TAD boundaries (blue), Peakachu chromatin loop boundaries (red), and SpectralTAD called TAD boundaries (green) on GM12878 and (D) on K562.

**Supplementary Figure 6. *preciseTAD* boundaries are more enriched for known molecular drivers of 3D chromatin, as compared with Arrowhead boundaries.** Enrichment heatmaps comparing the signal distribution ofCTCF, RAD21, SMC3, and ZNF143 around Arrowhead-called TAD boundaries vs. *preciseTAD*-predicted TAD boundaries for (A) GM12878 and (B) K562 cell lines.

**Supplementary Figure 7. *preciseTAD* boundaries are more enriched for known molecular drivers of 3D chromatin, as compared with Peakachu boundaries.** Enrichment heatmaps comparing the signal distribution ofCTCF, RAD21, SMC3, and ZNF143 around Peakachu chromatin loop boundaries vs. *preciseTAD*-predicted TAD boundaries for (A) GM12878 and (B) K562 cell lines.

**Supplementary Figure 8. *preciseTAD* boundaries are spatially closer to known molecular drivers of 3D chromatin.** Boxplots comparing the log2 genomic distance distributions from predicted and called boundaries to the nearest (A) GM12878-specific and (B) K562-specific CTCF, RAD21, SMC3, and ZNF143 transcription factor binding sites. p-values are derived from the Wilcoxon Rank Sum test.

**Supplementary Figure 9. The agreement between *preciseTAD*-predicted boundaries using Arrowhead- and Peakachu-trained models**. Venn diagrams of boundary overlaps using (A) GM12878 and (B) K562 data. Boundaries involving Arrowhead/Peakachu were flanked by 5 kb/10 kb, respectively.

**Supplementary Figure 10. Pre-trained *preciseTAD* models accurately predict boundaries on cell lines using annotation data only.** Venn diagrams and signal profile plots comparing flanked predicted boundaries using Arrowhead (A, B) and Peakachu (C, D) trained models. (A, C) train on GM12878 and predict on GM12878 (red, GM on GM) vs. train on K562 and predict on GM12878 (blue, K on GM), (B, D) train on K562 predict on K562 (red, K on K) vs. train on GM12878 predict on K562 (blue, GM on K). Boundaries involving Arrowhead/Peakachu were flanked by 5 kb/10 kb, respectively.
